---
title: "Using joins to create MLB tables"
author: "Group 9"
date: "M7 ICA1"
output: 
  pdf_document
    
---

<style type="text/css">
/* Title */
h1.title {
  color: #000088;
  font-size:60px;
  font-weight: bold;
}
/* Level 1 header */
h1 {
  color: #000088;
  font-weight: bold;
}
/* Level 2 header */
h2 {
  color: #CC0000;
  font-weight: bold;
}
/* Level 4 header */
h4 {
  color: #000000;
  font-weight: bold;
}
/* Table of contents */
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #CC0000;
    border-color: #337ab7;
}

</style> 

```{r setup, include=FALSE}
# global chunk options
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA,
                      fig.width=9, 
                      fig.height=7)
```

# Introduction

Today you will get practice merging data frames with inner and outer join
functions available in package `dplyr`. To get started, load packages
`tidyverse` and `Lahman`.

Package `Lahman` has numerous data frames about Major League Baseball.
Type `help(package = "Lahman")` in your Console to see everything available.

```{r packages}
library(tidyverse)
library(Lahman)
```

```{r hidden-packages, include=FALSE}
# library(extrafont)
# loadfonts(device = "win")
```

```{r theme-custom, include=FALSE}
theme_custom <- function() {
   theme_bw() +
   theme(#text = element_text(family = "Comic Sans MS", color = "#000088"),
         axis.title = element_text(size = 16), 
         title = element_text(size = 20),
         legend.title = element_text(size = 10),
         legend.text = element_text(size = 10),
         axis.text.x = element_text(size = 12),
         axis.text.y = element_text(size = 12),
         plot.caption = element_text(size = 10))
 }
```

# Data

You will work with data frames in package `Lahman`. When needed, utilize R's
help to get an understanding of what the variables are in a given data frame.
For example, `?Salaries` will provide a short description for each of the 
variables in data frame `Salaries`.

# Questions

## Question 1 

Select three data frames from package `Lahman`. Identify what variables are
in common between any pair of the three data frames, and identify what
variables are in common between all three data frames. What are the primary
keys for each data frame?

```{r}

names(Batting)
names(Pitching)
names(Fielding)
```
All 3 data frames have: playerID, yearID, stint, teamID, lgID, G 

Batting and Pitching share: HBP, SF, H
Batting and Fielding share: CS 
Pitching and Fielding share: GS

Batting primary key for the data frames is Player ID
Pitching primary key for the data frames is Player ID
Fielding primary key for the data frames is Player ID

## Question 2 

Use data frames `Managers` and `AwardsManagers` to reproduce the 
data frame given in the html output.


```{r}

full_join(Managers, AwardsManagers) %>% 
  filter(yearID==2016)
 
```


## Question 3

Use data frames `Managers`, `AwardsManagers`, and `Master` to reproduce the 
data frame given in the html output.


```{r}
mawards <- inner_join(Managers, AwardsManagers)
full_join(mawards, Master) %>%
select(yearID,teamID,  nameFirst, nameLast, awardID) %>%
  filter(yearID >= 2000)
  

```


## Question 4

Use data frames `Teams` and `Salaries` to reproduce the plot given in the html output. Bar colors
represent each team's primary color.

```{r}
  
Winners <- Teams%>%
  filter(WSWin == "Y")%>%
  select("teamID", "yearID")

winner.salaries <- Salaries%>%
  semi_join(Winners, by = c("teamID", "yearID"))

totals <- winner.salaries%>%
  filter(yearID > 1999)%>%
  group_by(teamID, yearID)%>%
  summarise(sal = sum(salary)/1000000)%>%
  arrange(yearID)

  ggplot(totals, aes(x = factor(yearID), y = sal, fill = factor(yearID))) +
  geom_bar(stat = "identity", position = "dodge") +
    theme_custom() +
  scale_x_discrete(labels = totals$teamID) +
    coord_flip() +
  scale_fill_manual(values = c("#0C2340", "#5F259F", "#003263", "#FF6600", "#BD3039", "#27251F", "#C41E3A", "#BD3039", "#E81828", "#0C2340", "#FD5A1E", "#C41E3A", "#FD5A1E", "#BD3039", "#FD5A1E", "#004687", "#0E3386")) +
    geom_text(aes(label = yearID))
```


## Question 5

Adjust your plot in Question 4 for inflation with year 2000 as baseline.
Comment on the differences between the plots.

Year | USD Value | Inflation Rate
-----|-----------|----------------
2000|	$1.00|	3.36%
2001|	$1.03|	2.85%
2002|	$1.04|	1.58%
2003|	$1.07|	2.28%
2004|	$1.10|	2.66%
2005|	$1.13|	3.39%
2006|	$1.17|	3.23%
2007|	$1.20|	2.85%
2008|	$1.25|	3.84%
2009|	$1.25|	-0.36%
2010|	$1.27|	1.64%
2011|	$1.31|	3.16%
2012|	$1.33|	2.07%
2013|	$1.35|	1.46%
2014|	$1.37|	1.62%
2015|	$1.38|	0.12%
2016|	$1.39|	1.26%

```{r}
usd = c(1.00, 1.03, 1.04, 1.07, 1.10, 1.13, 1.17, 1.20, 1.25, 1.25, 1.27, 1.31, 1.33, 1.35, 1.37, 1.38, 1.39)
length(usd)
length(Winners$sal*usd)
n.salary = Winners$sal*usd

Winners%>%
  ggplot(mapping = aes(y = n.salary, x = as.factor(yearID), fill = teamID)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = totals$teamID) +
  scale_fill_manual(values = c("#0C2340", "#5F259F", "#003263", "#FF6600", "#BD3039", "#27251F", "#C41E3A", "#BD3039", "#E81828", "#0C2340", "#FD5A1E", "#C41E3A", "#FD5A1E", "#BD3039", "#FD5A1E", "#004687", "#0E3386")) +
  coord_flip() +
  theme_custom() +
  geom_text(aes(label = yearID), color = "white", position = position_stack(vjust = 0.9)) +
  labs(title = "Payroll of world series winners", x = "World Series Winners", y = "Salary in millions")
```

# References {#References-link}

1. Lahman, S. (2017) Lahman's Baseball Database, 1871-2016.

2. RStudio Cheatsheets