---
title: "Tidy data"
author: "Kaitlyn Watson-Group 9"
date: "Module 2 ICA 3 "
output: 
  pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)
```

# Introduction

Messy data is common. A large portion of your time may be spent cleaning,
parsing, and organizing a data set. Tidy data is the ultimate goal. Four
functions available in the `tidyr` package will help make this process easier:

- `gather()`

- `spread()`

- `separate()`

- `unite()`

Package `tidyr` is automatically loaded when you load `tidyverse`.

```{r packages, echo=FALSE}
library(tidyverse)
```

Open the file `gapminder-raw.csv` to see what it contains and its format.
Read in `gapminder-raw.csv` to R and save it as an object called `gapminder1`. 
Check a few of the rows below to make sure your data was read in
properly.

```{r read-data, echo=FALSE}
gapminder1<-read.csv("gapminder-raw.csv")
gapminder1
```
**Comment on why the above data frame is not tidy?**
Column names are unclear and the life expectancy collumn is above the years listed. Each column represents one year. Additionally, there are many areas with N/A present throughout the entire data set. 

# Tidy data { - .tabset .tabset-fade .tabset-pills}

You will now tidy `gapminder1` by using a series of functions in the `tidyr` 
package.

In each step, examine the resulting data frame, and attempt to produce code that 
generates the resulting data frame. Carefully examine the variable names, types,
and first few rows.

## Step 1: Wide to long { - .tabset .tabset-fade .tabset-pills}


```{r wide to long, echo=FALSE}

gapminder1<-gather(gapminder1, key=year, value=life.exp, -Life.expectancy)
gapminder1
```
Years are now represented in the rows rather than in the columns. And life expectancy has the correct values in the designated column.  
### Hints

- Function `gather()` takes multiple columns and collapses into key-value pairs, 
  duplicating all other columns as needed. You use `gather()` when you 
  notice that you have columns that are not variables.
  
- Function `gather()` will transform a data frame from wide to long format.

- You want to gather all but the first column of `gapminder1`.

- Run each line of the code below in your console for a small example.



## Stpe 2: Variable names { - .tabset .tabset-fade .tabset-pills}


```{r add variable names}
names(gapminder1)[1]<-"country"
gapminder1
```


## Step 3: Fixing the years { - .tabset .tabset-fade .tabset-pills}


```{r tidying the data, echo=FALSE}
gapminder1<-separate(gapminder1, "year", into=c("X", "year"), "X")
gapminder1<-gapminder1[ -c(2) ] 
gapminder1$year<-as.integer(gapminder1$year)
gapminder1
```


# Visualizations using base R functions { - .tabset .tabset-fade .tabset-pills}

Recreate plots 1 and 2. Try to create the plot without looking at the hints, and **comment on any interesting trends/relationships you observe**.

## Plot 1 { - .tabset .tabset-fade .tabset-pills}

### Plot {-}

```{r plot1, echo=FALSE, cache=TRUE}
US<-subset(gapminder1,gapminder1$country=="United States")

plot(US$year, US$life.exp, type = "l", lwd=3, Main="US Mean Life Expectancy Increase", sub="Large drops during the Civil War and WW1", col="red") 
abline(v=1863)
abline(v=1918)
```

### Hints {-}

- Use `subset()` to filter `gapminder` for United States

- `plot(x = year, y = life.exp, type="l",...)`

- Use `lwd`, `main`,`sub`,`xlab` and `ylab` in the `plot()`

- `abline(v=1863)`

## Plot 2 { - .tabset .tabset-fade .tabset-pills}

### Plot 

```{r plot2, echo=FALSE, cache=TRUE}
Countries <- c(gapminder1, gapminder1$country == "China", "India", "Indonesia", "United States", "Brazil")

plot(gapminder1$year[gapminder1$country == "China"],gapminder1$life.exp[gapminder1$country == "China"], type = "l", xlab = "Year", ylab = "Life Expectancy", lwd = 2, ylim = c(0,90))
lines(gapminder1$year[gapminder1$country=="India"], gapminder1$life.exp[gapminder1$country=="India"], type = "l", col = "blue", lwd=2)
lines(gapminder1$year[gapminder1$country=="Indonesia"], gapminder1$life.exp[gapminder1$country=="Indonesia"], type = "l", col = "red", lwd=2)
lines(gapminder1$year[gapminder1$country=="United States"], gapminder1$life.exp[gapminder1$country=="United States"], type = "l", col = "pink", lwd=2)
lines(gapminder1$year[gapminder1$country=="Brazil"], gapminder1$life.exp[gapminder1$country=="Brazil"], type = "l", col = "orange", lwd=2)

legend("topleft", legend = c("China", "India", "Indonesia", "US", "Brazil"), col = c("black", "blue", "red", "pink", "orange"), lty = 1, cex = 0.5)
```
The United States has a considerably higher life expectancy as the years progress; however, the common trend between all countries is increasing. 
### Hints 

- Use `subset()` to filter `gapminder` for 
  `c("China", "India", "Indonesia", "United States", "Brazil")`
  
- `plot(year[country=="China"], life.exp[country=="China"],...)` for "China".

- Use `lines(year[country=="India"], life.exp[country=="India"], type = "l", col = "red", lwd=2)` for each additional country.

- legend("topleft", legend = c("China", "India", "Indonesia", "US", "Brazil"), col = c("black", "red", "green","orange","purple"),
       lty = 1, cex = 0.5)



# References

1. https://www.gapminder.org/
